Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-06-15 020-simulation-algorithm.Rmd
migrate to html based preview

### Inputs
(1) Marginal characteristics including the same of distributional family and parameter values, denotes as $F_i$ for marginal component random variable $Y_i$ for $i=1,\ldots,d$.
(2) A target (specified) Spearman's correlation matrix ${\bf R_{Spearman}}$ with
each $\rho_{Spearman}$ within the Frechet limits. Alternatively, the rescaled
Spearman's correlation (see Equation \@ref(eq:convertKendallDiscrete) ) can be provided.
(3) An error tolerance $\epsilon$ for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.

### Algorithm
(i) Compute Spearman's $\rho_{rs}$ (see Equation
\@ref(eq:convertKendallDiscrete)) from the specified $\rho_{spearman}$ for each
pair of marginal random variables $(Y_i,Y_{i^\prime})$ with $i \neq i^\prime$
when either marginal is discrete. In such large scale computations, it may be
that numerically $\rho_{rs}$ is be larger/smaller than the Frechet bounds (or
even $\pm 1$). To guard against this, set $\rho_{rs} = min( \rho_{rs}, M)$,
where $M$ is the upper Frechet bound (see [Section Introduction]) and similarly set $\rho_{rs} = max( \rho_{rs}, W)$, where $W$ is the lower Frechet bound for this pair of distributions. Gather these $\rho_{rs} \lq s$ into a new input correlation matrix ${\bf R_{rs}}$.  
(ii) Convert ${\bf R_{Spearman}}$ into ${\bf R_{Pearson}}$ via $\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)$.   
(iii) Finally, ensure that ${\bf R_{Pearson}}$ is positive definite, by finding the nearest positive definite correlation matrix ${\bf R}$ --- using the `R` routine `nearPD` in the `Matrix` package --- within an error tolerance of $\epsilon$.  
(iv) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
(v) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(vi) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

>>>>

### Algorithm 1: Matching Pearson approximately

We aim to simulate multivariate data with a Pearson correlation
matrix ${\bf R}$.

(i) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})$;  
(ii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iii) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

* Where $F_i^{-1}$ is the inverse CDF for the $i^{th}$ gene's marginal
distribution.  
* And ${\bf R}$ is the target Pearson correlation matrix.  
* Pearson's correlation is **not** invariant under the montonically increasing
  transformations in steps 3, 4.  
* So the resulting simulated random vectors will have the **exact** margins but **approximate** ${\bf R}$.

### Algorithm 2: Matching Spearman's exactly

(i) Convert ${\bf R}_{Spearman}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} =
  2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf} R_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Spearman's $\rho$ relates to the difference in the ranks of pairs of values.
- We take advantage of a closed form correspondence between Pearson and Spearman
  for *normal random variables* in step 1 [@BF17].
- Since Spearman's $\rho$ is invariant under the monotonically increasing
  transformations in steps 3 and 4, the resulting simulated random vectors will
  have the **exact** margins and **exact** Spearman correlation (up to Monte Carlo/numeric error).

### Algorithm 3: Matching Kendall's exactly

(i) Convert ${\bf R}_{Kendall}$ into ${\bf R}_{Pearson}$ via $\rho_{Pearson} = sin \left( \tau \times \frac{\pi}{2} \right)$;  
(ii) Generate ${\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R}_{Pearson})$;  
(iii) Transform ${\bf X}$ to ${\bf U} = (U_1, \ldots,  U_d)$ viz $U_i=\Phi(X_i)$, $i=1, \ldots, d$;  
(iv) Return ${\bf Y}  = (Y_1, \ldots,  Y_d)$, where $Y_i=F_i^{-1}(U_i)$, $i=1, \ldots, d$;

- Kendall's $\tau$ is the difference in the probabilities of concordant and discordant pairs.
- We use the 1-1 relationship between Pearson and Kendall's $\tau$ for *normal
  rvs* in step 1.
- Since Kendall's $\rho$ is invariant under the montonically increasing
  transformations in steps 3 and 4,  
 the resulting simulated random vectors will have the **exact** margins and
 **exact** Kendall's $\tau$  (up to Monte Carlo/numeric error).

## Algorithm notes

- Discrete margins pose some difficulties and matching *unadjusted* correlation
  measures is not exact (empirically approximate).
- Exact Pearson solutions are possible but require computing $d \choose 2$
   double integrals $EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)),
   F_j^{-1}(\Phi(z_j))dy_idy_j$. We have yet to accelerate this process.
- We GPU accelerate whenever possible during our proposed High-Performance NORTA
  algorithm.
- One thorny problem when dealing with high-dimensional correlation matrices is
they can become non-positive definite through either during estimation or
bivariate transformations.
- When converting nonparametric correlations to Pearson (first step above) the
  resultant correlation may not be positive definite (PD).
- In that case, we replace ${\bf R}_{Pearson}$ with a "close" PD matrix ${\bf \tilde{R}}_{Pearson}$.
- In practice, this loss in accuracy typically has little impact on performance,
  but the algorithm needs acceleration (`Matrix::nearPD()` violates property S1).

test workflow again

### 2020-06-15 030-bigsimr-package.Rmd
migrate to html based preview

```{r setup, eval=TRUE}
## Using conda
reticulate::use_condaenv("py37")
library(bigsimr)
## help(package = 'bigsimr')
```

@@

```{r}
margins = list(
  list("norm", mean = 3.14, sd = 0.1),
  list("beta", shape1 = 1, shape2 = 4),
  list("nbinom", size = 10, prob = 0.75)
)
```

@@

```{r}
x = rvec(10, rho = rho, params = margins, type = "pearson")
```

On my machine, there is no dedicated GPU, so I would see the following warning message once per session.

@@

```{r, fig.width=7}
x = rvec(10000, rho = rho, params = margins, type = "pearson")

@@

```{r}
cor(x)
```

We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation. This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions. Another possibility is that the copula algorithm is biased and needs correction.

>>>>

```{r setup, eval = TRUE}
library(bigsimr)
# Reticulate needs to be able to find the python binary with `jax` installed
reticulate::use_condaenv("bigsimr-cpu")
```

@@

```{r}
margins = alist(
  qnorm(mean = 3.14, sd = 0.1),
  qbeta(shape1 = 1, shape2 = 4),
  qnbinom(size = 10, prob = 0.75)
)
```

@@

```{r}
x <- rvec(10, rho = rho, margins = margins, type = "spearman")
```

@@

```{r, fig.width=7}
x <- rvec(100000, rho = rho, margins = margins, type = "spearman")

@@

```{r}
cor(x, method = 'spearman')
```

We can see that even with 100,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation. This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions. Another possibility is that the copula algorithm is biased and needs correction.

### 2020-06-15 050-some-applications.Rmd
migrate to html based preview

@@

```{r brca, echo=TRUE, eval=FALSE}
## full workflow simulating RNA-seq data

@@

```{r figNBratio, echo = FALSE, eval=FALSE, message=FALSE, warning=FALSE, fig.path='fig/plot-', out.width= '75%', fig.align='center', fig.ext='pdf', fig.cap="The genes display extreme overdispersion. Using group-specific sample means and variances for each of the d=4777 genes, we display the distributions oflog of variance to mean ratio (log(sigma2/mu))."}
## http://www.sthda.com/english/wiki/ggplot2-histogram-easy-histogram-graph-with-ggplot2-r-package
## devtools::install_github("kassambara/easyGgplot2")
library(easyGgplot2)
## p0 <- ggplot(data = nb_dat, aes(x = log(nb_var / nb_mean), y = stat(density), fill=group))
## p0 + geom_histogram(bins = 30) + labs(x="log of variance to mean ratio") + theme_bw()
nb_dat$logRatio <- log(nb_dat$nb_var / nb_dat$nb_mean)
ggplot2.histogram(data=nb_dat, xName='logRatio',
    groupName='group', legendPosition="right",
    alpha=0.5, addDensity=TRUE,
    addMeanLine=TRUE, meanLineColor="white", meanLineSize=1.5)
```

@@

- Kendall's $\tau$ was estimated efficiently using the `pcaPP::cor.fk` function in `R`.

## Ultra High-Dimensional simulation-based correlation hypothesis testing

```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
## full workflow simulating RNA-seq data

>>>>

```{r LoadLib050, echo = FALSE, eval = TRUE, message=FALSE, warning=FALSE, results = "hide"}
library(ggplot2)
library(tidyverse)
library(knitr)
library(dplyr)
## devtools::install_github( "adknudson/bigsimr", ref = 'develop')
library(bigsimr)
reticulate::use_condaenv("py37")
CORES <- parallel::detectCores() - 1
set.seed(06082020)
## devtools::install_github("ggobi/GGally")
## devtools::install_github( "adknudson/bigsimr")
## A conda environment is also activitied (see README)
library(bigsimr)
library(GGally)
```

@@

```{r processBRCA, echo=TRUE, eval=TRUE, cache=TRUE}
## full workflow simulating RNA-seq data
allDat <- readRDS( file = "~/Downloads/complete_processed_tcga2stat_RNASeq2_with_clinical.rds" )
lastClinical <- which( names(allDat) == 'tumorsize' )
brca <- allDat[ allDat$disease == "BRCA", (lastClinical+1):ncol(allDat) ]
## remove naively the RSEM adjustment
brca <- round(brca, 0)
ncol(brca) ## num of genes 20501
## compute the median expression! avoid the 0s
brcaMedian <- apply(brca, 2, median)
## retain top (1-probs)*100% highest expressing genes for illustration
myProb <- 0.95 
cutPoint <- quantile( x = brcaMedian, probs = myProb )
## genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
genesToKeep <- names( brcaMedian ) [ which(brcaMedian >= cutPoint) ]
brca <- brca[ , genesToKeep ]
## ncol(brca) / 20501
(d <- length(genesToKeep))
```

@@

```{r figNBratio, echo = TRUE, eval=TRUE, message=FALSE, warning=FALSE, out.width= '75%', fig.align='center', fig.cap='The genes display extreme overdispersion. Using group-specific sample means and variances for each of the d=`r d` genes, we display the distributions oflog of variance to mean ratio.'}

logVarOverMean <- function( x ) {
    log ( var(x) / mean (x) )
}
## check whether NB makes sense

summaryBRCA <- brca %>% 
    summarize_all(  list(~ logVarOverMean( . ) ) )
summaryBRCA <- as.data.frame( t(as.data.frame( summaryBRCA )) )
names(summaryBRCA) <- "logVarOverMean"
## sum(summaryBRCA$logVarOverMean < 0)  ## no underdispersed genes
ggplot(data = summaryBRCA, mapping = aes(x = logVarOverMean)) +
  geom_histogram(color = "white", bins = 20)
```

@@

```{r estRhoBRCA, echo=TRUE, eval=TRUE}
## full workflow simulating RNA-seq data

## 1. Estimate Spearman's correlation on the count data
corType <- 'spearman'
system.time( rho <- bigsimr::fastCor( brca, method = corType ) )
## Describe the correlations
```

```{r estMargins, echo = TRUE, eval = TRUE, cache=TRUE}
## 2. Estimate NegBin parameters using Method of Moments
estimateNegBinMoM <- function(tmpGene, minP = 1e-5) {
    tmpMean <- mean(tmpGene)
    tmpVar <- var(tmpGene)
    ## relate to nbinom parameters
    ## See ?rbinom for details.
    p <- tmpMean / tmpVar
    if (p < minP) {p <- minP } ## maybe add noise here
    n <- ( tmpMean * p) / ( 1  - p )  
    ## format for bigsimr margins
    return( list("nbinom", size = n, prob = p) )
}
brcaMargins <- apply( unname(as.matrix(brca)), 2, estimateNegBinMoM )
## describe the margins
## check for too small prob
## head( sort( unlist( lapply( brcaMargins, function(x) {x$prob} ) ) ) )
## head( sort( unlist( lapply( brcaMargins, function(x) {x$size} ) ) ) )
```

```{r simBRCA, echo=TRUE, eval = FALSE, cache=TRUE}
## 3. Generate the simulated samples
## N <- 10
## system.time( simBRCA <- rvec(N, rho = rho[1:2, 1:2], params = brcaMargins[1:2], type = corType, adjustForDiscrete = FALSE) )
N <- nrow(brca)
system.time( simBRCA <- rvec(N, rho = rho, params = brcaMargins, cores = CORES, type = corType, adjustForDiscrete = FALSE) )
simBRCA[1:2, 1:2 ]

```{r evalBRCA, echo=TRUE, eval = FALSE, cache=TRUE}

## Describe and plot real data
GGally::ggpairs(data = as.data.frame(brca[ ,1:3] ) )

## Describe and plot simulations
GGally::ggpairs(data = as.data.frame(simBRCA[ ,1:3] ) )

## check 1st moments
trueMu <- colMeans(brca)
simMu <- colMeans(simBRCA)
qplot( x = trueMu, y = simMu ) + geom_abline(slope = 1, intercept = 0)

## check 2nd moment
trueVar <- apply( brca, 2, var)
simVar <- apply( simBRCA, 2, var)
qplot( x = trueVar, y = simVar ) + geom_abline(slope = 1, intercept = 0)

## check correlation
system.time( simRho <- bigsimr::fastCor( simBRCA, method = corType ) )
rho[1:5, 1:5]
simRho[1:5, 1:5]
trueRho <- rho[lower.tri(rho)]
simRho <- simRho[lower.tri(simRho)]
qplot( x = trueRho, y = simRho ) + geom_abline(slope = 1, intercept = 0)
```

## Ultra High-Dimensional simulation-based correlation hypothesis testing under

Since many of the correlations are near zero, one could ask whether it is
possible that

```{r brcaSigmaTesting, echo=TRUE, eval=FALSE}
